---
title: "Project_Text"
author: Leslie Cervantes Rivera & Kevin Le
format: pdf
editor: visual
---

# Background

The Infant Mortality and Fertility Rates Dataset provides data on infant mortality rates (measured as the number of infant deaths per 1,000 live births) and fertility rates (measured as the number of births per 1,000 women per year) across U.S. states. The Infant Mortality Dataset covers the years 2003 to 2023, while the Fertility Rates Dataset spans 2016 to 2023.

Infant mortality refers to the death of baby that occurs before their first birthday. It is a key indicator of a population's overall health, reflecting the social, economic, and healthcare conditions within a state. High infant mortality rates could indicate insufficient healthcare access or inequalities in medical services, particularly in low-income and rural areas. Over the years, improvements in screening and treatment for illnesses, better obstetric management, and neonatal care have contributed in the declining of infant mortality rates, but disparities still exist within state and demographic groups. (<https://www.vcuhealth.org/news/infant-mortality-rates-declining-but-sudden-unexpected-infant-death-is-on-the-rise/#:~:text=Wolf%20attributes%20declining%20overall%20infant,media%20on%20infant%20sleep%20practices.>) (<https://www.nichd.nih.gov/health/topics/infant-mortality/topicinfo>)

Fertility rate represents the total number of children a woman has during her reproductive years. It plays a crucial role in population growth and demographical planning, influencing services such as education, healthcare, and workforce development. From 2007 to 2022, the fertility rate has dropped by about 19%, influenced by the health of the economy, social, health, and historical events. These changes have long-term implications, as it leads to an aging population, smaller workforce, and economic strain on government budgets. (<https://www.britannica.com/topic/fertility-rate>) (<https://www.northwell.edu/news/the-latest/us-fertility-rate-decline-impact#:~:text=After%20a%20few%20decades%20of,historical%20events%E2%80%94affect%20family%20sizes.>)

The Maternal Mortality dataset provides national-level provisional counts of deaths among women who die during pregnancy or within 42 days of the termination of pregnancy, but not from accidental or incidental causes. These deaths are based on death records that have been received and processed by the National Center for Health Statistics (NCHS) as of a specified cutoff date. The dataset includes deaths among U.S. residents that occurred within the 50 states and the District of Columbia. Maternal mortality rates are calculated as deaths per 100,000 live births.

The Population Demographics dataset provides data on the years 2016 to 2023, including total population of sex and age, marital status, educational attainment, SNAP, and health insurance coverage based on ethnicity. 

## Data Sources

### Birth Rates and Infant Mortality Datasets 

The datasets come from the Centers for Disease Control and Prevention (CDC), collected from anywhere a person receives healthcare. This may lead to limitations as it is up to the city, county, and state to decide what information is collected, and how and when it can be shared by the CDC. (<https://archive.cdc.gov/www_cdc_gov/surveillance/data-modernization/basics/where_does_our_data_come_from.html>)

The NCHS obtains maternal death records from state vital registration offices through the Vital Statistics Cooperative Program. The estimates of maternal mortalities are derived from final mortality data for the years 2019 to 2021, and provisional data for the years 2022 to 2024. Provisional data reflects counts based on reports that are still being processed, while finalized data represents more complete counts of maternal deaths after thorough review and analysis. (Provisional Maternal Death Counts) https://www.cdc.gov/nchs/nvss/vsrr/provisional-maternal-deaths.htm

median income -\> 2020 reasoning <https://www.census.gov/data/developers/data-sets/acs-1year.2020.html#list-tab-843855098>

### Population Demographics

The dataset comes from the United States Census Bureau under the category American Community Survey named Selected Population Profile in the United States. The Census Bureau collects their data from respondents directly through censuses and surveys. Primary sources come from administrative data.  


### Infant mortality

| Variable Name | Meaning |
|----|----|
| `bacode` | Biannual code. 1 means data aggregated from January 1 through June 30, 2 means data from July 1 through December 31. |
| `births_nhblack` | Number of births to non-Hispanic Black mothers |
| `births_nhwhite` | Number of births to non-Hispanic White mothers |
| `births_hisp` | Number of births to Hispanic mothers of any race |
| `births_otherraceeth` | Number of births to mothers of other races/ethnicities |
| `births_total` | Total number of births across all racial/ethnic groups |
| `deaths_hisp` | Number of deaths among Hispanic population of any race |
| `deaths_con` | Number of congenital-related deaths (present at birth) |
| `deaths_noncon` | Number of non-congenital deaths |
| `deaths_neo` | Number of neonatal deaths (within first 28 days of life) |

### Fertility

| Variable Name         | Meaning                                            |
|-----------------------|----------------------------------------------------|
| `births_age1524`      | Births to mothers aged 15-24                       |
| `births_age2534`      | Births to mothers aged 25-34                       |
| `births_age3544`      | Births to mothers aged 35-44                       |
| `pop_age1524`         | Population aged 15-24 (women age 15-54 population) |
| `pop_age2534`         | Population aged 25-34 (women age 15-54 population) |
| `pop_age3544`         | Population aged 35-44 (women age 15-54 population) |
| `births_nhblack`      | Births to non-Hispanic Black mothers               |
| `births_nhwhite`      | Births to non-Hispanic White mothers               |
| `births_hisp`         | Births to Hispanic mothers                         |
| `births_otherraceeth` | Births to mothers of other races/ethnicities       |
| `births_total`        | Total births across all categories                 |

#### Education

| Variable Name | Meaning |
|----|----|
| `births_nohs` | Births to mothers without high school education |
| `births_hs` | Births to mothers with high school education |
| `births_somecoll` | Births to mothers with some college education |
| `births_coll` | Births to mothers with college degree |
| `pop_nohs` | Population without high school education (women age 15-54 population) |
| `pop_hs` | Population with high school education (women age 15-54 population) |
| `pop_somecoll` | Population with some college education (women age 15-54 population) |
| `pop_coll` | Population with college degree (women age 15-54 population) |

#### Status

| Variable Name | Meaning |
|----|----|
| `births_married` | Births to married mothers |
| `births_unmarried` | Births to unmarried mothers |
| `pop_married` | Married population (women age 15-54 population) |
| `pop_unmarried` | Unmarried population (women age 15-54 population) |
| `births_medicaid` | Births covered by Medicaid |
| `births_nonmedicaid` | Births not covered by Medicaid |
| `pop_medicaid` | Population covered by Medicaid (women age 15-54 population) |
| `pop_nonmedicaid` | Population not covered by Medicaid (women age 15-54 population) |

## Question of Interest


## Exploratory Data Analysis



```{r}
library(ggplot2)
library(janitor)
maternal_mort1 <- maternal_mort |>
  pivot_wider(names_from= Group, values_from=Subgroup)

maternal_mort1 <- maternal_mort1 %>% 
  clean_names()

maternal_byAge <- maternal_mort1 |> filter(!is.na(age))
maternal_byRace <- maternal_mort1 |> filter(!is.na(race_and_hispanic_origin)) |>
  rename("RaceEth"=race_and_hispanic_origin)

head(maternal_byAge)
head(maternal_byRace)
okabe <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#D55E00", "#CC79A7")

maternal_byRace |> ggplot(aes(x=year_of_death, y=maternal_mortality_rate, color=RaceEth)) +
  geom_point()
maternal_byRace |> ggplot(aes(x=live_births, y=maternal_deaths, color=as.factor(year_of_death))) +
  geom_point() + 
  scale_color_manual(values=okabe)

```

```{r}
fertility_data |> select(year, policy_on) |> filter(policy_on==1) |> summarize(n=n(), .by=year )
infant_data |> select(year, policy_on) |> filter(policy_on==1) |> summarize(n=n(), .by=year )
```

The `echo: false` option disables the printing of code (only output is displayed).

**Data description:** write up a short summary of the dataset you chose to work with following the conventions introduced in previous assignments. Cover the sampling if applicable and data semantics, but focus on providing high-level context and not technical details; don’t report preprocessing steps or describe tabular layouts, etc.

**Question of interest:** motivate and formulate your question; explain what a satisfactory answer might look like.

**Data analysis:** provide a walkthrough with commentary of the steps you took to investigate and answer the question. This section can and should include code cells and text cells, but you should try to focus on presenting the analysis clearly by organizing cells according to the high-level steps in your analysis so that it is easy to skim. For example, if you fit a regression model, include formulating the explanatory variable matrix and response, fitting the model, extracting coefficients, and perhaps even visualization all in one cell; don’t separate these into 5-6 substeps.

**Summary of findings:** answer your question by interpreting the results of your analysis, referring back as appropriate. This can be a short paragraph or a bulleted list.

# Background

The Infant Mortality and Fertility Rates Dataset provides data on infant mortality rates (measured as the number of infant deaths per 1,000 live births) and fertility rates (measured as the number of births per 1,000 women per year) across U.S. states. The Infant Mortality Dataset covers the years 2003 to 2023, while the Fertility Rates Dataset spans 2016 to 2023.

Infant mortality refers to the death of baby that occurs before their first birthday. It is a key indicator of a population's overall health, reflecting the social, economic, and healthcare conditions within a state. High infant mortality rates could indicate insufficient healthcare access or inequalities in medical services, particularly in low-income and rural areas. Over the years, improvements in screening and treatment for illnesses, better obstetric management, and neonatal care have contributed in the declining of infant mortality rates, but disparities still exist within state and demographic groups. (<https://www.vcuhealth.org/news/infant-mortality-rates-declining-but-sudden-unexpected-infant-death-is-on-the-rise/#:~:text=Wolf%20attributes%20declining%20overall%20infant,media%20on%20infant%20sleep%20practices.>) (<https://www.nichd.nih.gov/health/topics/infant-mortality/topicinfo>) (<https://www.cdc.gov/nchs/data_access/vitalstatsonline.htm>)

Fertility rate represents the total number of children a woman has during her reproductive years. It plays a crucial role in population growth and demographical planning, influencing services such as education, healthcare, and workforce development. From 2007 to 2022, the fertility rate has dropped by about 19%, influenced by the health of the economy, social, health, and historical events. These changes have long-term implications, as it leads to an aging population, smaller workforce, and economic strain on government budgets. (<https://www.britannica.com/topic/fertility-rate>) (<https://www.northwell.edu/news/the-latest/us-fertility-rate-decline-impact#:~:text=After%20a%20few%20decades%20of,historical%20events%E2%80%94affect%20family%20sizes.>)

Maternal mortality is a leading cause of death and disability for adult women worldwide, responsible for an estimated 289,000 deaths in 2013 \[1\]. It represents true excess burden of disease since the overwhelming majority of maternal deaths are due to preventable causes; and could be treated with well-understood interventions that have long been available in the global North. If the infant survives birth but the mother does not, the resulting lack of nutritional support from breastfeeding leaves the baby vulnerable to malnutrition, which can itself be fatal or may increase the risk of disease or death from infection \[6–8\]. Older siblings also may suffer in many ways without maternal care: among orphans, the risk of child labor \[9, 10\], poor learning outcomes and lower educational attainment \[11\], and disrupted living arrangements \[12\] can impose trauma that has detrimental impacts on health and well-being.

The Maternal Mortality dataset provides national-level provisional counts of deaths among women who die during pregnancy or within 42 days of the termination of pregnancy, but not from accidental or incidental causes. These deaths are based on death records that have been received and processed by the National Center for Health Statistics (NCHS) as of a specified cutoff date. The dataset includes deaths among U.S. residents that occurred within the 50 states and the District of Columbia. Maternal mortality rates are calculated as deaths per 100,000 live births.

## Data Sources

The datasets come from the Centers for Disease Control and Prevention (CDC), collected from anywhere a person receives healthcare. This may lead to limitations as it is up to the city, county, and state to decide what information is collected, and how and when it can be shared by the CDC. (<https://archive.cdc.gov/www_cdc_gov/surveillance/data-modernization/basics/where_does_our_data_come_from.html>)

The CDC uses The Pregnancy Risk Assessment Monitoring System (PRAMS) collects data on attitudes and experiences before, during, and after pregnancy. PRAMS data can be used to measure progress toward goals in improving the health of mothers and infants.  

Each month at the jurisdiction level, a stratified sample is randomly drawn from the current birth certificate file. For each of these monthly samples, or "batches," a series of mailings begins 2 to 4 months after delivery. This is to allow new mothers time to respond to questions about the early postpartum period. With all mailings, an invitation is sent to complete the survey by web.

The NCHS obtains maternal death records from state vital registration offices through the Vital Statistics Cooperative Program. The estimates of maternal mortalities are derived from final mortality data for the years 2019 to 2021, and provisional data for the years 2022 to 2024. Provisional data reflects counts based on reports that are still being processed, while finalized data represents more complete counts of maternal deaths after thorough review and analysis. ([Provisional Maternal Death Counts](https://www.cdc.gov/nchs/nvss/vsrr/provisional-maternal-deaths.htm))

```{r}

```


